---
title: "Seed Classification using Decision Trees"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(123)
```

## **1. Load Required Libraries**
```{r}
library(ggplot2)
library(rpart)
library(rpart.plot)
library(caret)
library(gridExtra)
```

## **2. Load and Inspect Data**
```{r}
# Read the dataset
seeds <- read.csv("seeds.csv", header = TRUE)

# Convert Class to factor
seeds$Class <- factor(seeds$Class)

# Display dataset structure
str(seeds)

# Summary statistics
summary(seeds)

# Check for missing values
print(colSums(is.na(seeds)))
```

## **3. Handle Missing Values**
```{r}
# Impute missing values (replace with median)
seeds$Eccentricity[is.na(seeds$Eccentricity)] <- median(seeds$Eccentricity, na.rm = TRUE)
seeds$Solidity[is.na(seeds$Solidity)] <- median(seeds$Solidity, na.rm = TRUE)

# Verify missing values are handled
print(colSums(is.na(seeds)))
```

## **4. Split Data into Training & Testing Sets**
```{r}
# 80% Training, 20% Testing
trainIndex <- createDataPartition(seeds$Class, p = 0.8, list = FALSE)
trainData <- seeds[trainIndex, ]
testData <- seeds[-trainIndex, ]
```

## **5. Build Decision Tree Model**
```{r}
# Default Decision Tree
tree_default <- rpart(Class ~ ., data = trainData, method = "class")

# Plot the tree
rpart.plot(tree_default, main = "Default Decision Tree")
```

## **6. Fully Grown Decision Tree**
```{r}
# Fully grown tree (no stopping criteria)
tree_full <- rpart(Class ~ ., data = trainData, method = "class", control = rpart.control(minsplit = 2, cp = 0))

# Visualize the fully grown tree
rpart.plot(tree_full, main = "Fully Grown Decision Tree")
```

## **7. Prune the Decision Tree**
```{r}
# Check complexity parameter (cp)
printcp(tree_full)

# Prune using cp = 0.005
tree_pruned <- prune(tree_full, cp = 0.005)

# Visualize the pruned tree
rpart.plot(tree_pruned, main = "Pruned Decision Tree")
```

## **8. Model Evaluation - Confusion Matrix**
```{r}
# Predictions on test set
pred_pruned <- predict(tree_pruned, testData, type = "class")

# Convert actual and predicted values to factors with matching levels
actual <- factor(testData$Class)
predicted <- factor(as.character(pred_pruned), levels = levels(actual))

# Compute confusion matrix
conf_matrix <- confusionMatrix(predicted, actual)
print(conf_matrix)
```

## **9. Accuracy & Model Performance**
```{r}
# Extract Accuracy and Kappa
accuracy <- conf_matrix$overall["Accuracy"]
kappa <- conf_matrix$overall["Kappa"]

# Print results
cat("**Model Accuracy:**", round(accuracy * 100, 2), "%\n")
cat("**Kappa Statistic:**", round(kappa, 3), "\n")
```

## **10. Conclusion**
The decision tree model achieves **`r round(accuracy * 100, 2)`%** accuracy with a **Kappa of `r round(kappa, 3)`**. If interpretability is a priority, this model is a great choice. If higher accuracy is needed, other models like **Random Forest or SVM** may be explored.


